---
title: "Explore financial contribution of presidential election of California"
author: "Bordé Sándor"
date: ''
output: html_document
---

```{r global_options, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
```
I chose the *Financial contribution of presidential election* dataset (California). I have a little bit more than 1.3 million entry and 19 variables.
## Get to know the dataset

```{r}
dataSet <- read.csv('support_CA_col.csv', row.names = NULL)
dim(dataSet)
```

```{r}
str(dataSet)
```
```{r}
summary(dataSet)
```
```{r}
names(dataSet)
```

I have 1304346 observations and 18 variables from which 16 was factor variable.
## Univariate plots
```{r}
edges = seq(-11000, 11000, 1000)
hist(dataSet$contb_receipt_amt, breaks = edges)
```

```{r}
edges = seq(-11000, 11000, 200)
ggplot(dataSet, aes(contb_receipt_amt)) + 
  geom_histogram(breaks = edges)
```
```{r}
boxplot(dataSet$contb_receipt_amt)
```
```{r}
library(moments)
kurtosis(dataSet$contb_receipt_amt)
```
From these plots we can see that the contribution amount follows a narrow, non-normal distribution. It can be quantified by measuring kurtosis which is 64.19. As the kurtosis become larger, the shape of the distribution become sharper. That means that most of the values fall within a relatively narrow range compared to the whole range which is very wide.
The 1sq quartile is 15, the 3rd quartile is 88 therefor the IQR is 73, but the full data range is 21300. I am curious about the contribution of individual (I mean, average people) I made a plot with restricted range.
```{r}
restrictedSet <- subset(dataSet, dataSet$contb_receipt_amt>=0 & dataSet$contb_receipt_amt<=200)
edges = seq(0, 200, 5)
hist(restrictedSet$contb_receipt_amt, edges)
```
I found it interesting, that the distribution of the restricted set almost a long-tail distribution, but it has some "spikes" at some values. I plotted a new histogram with a finer bin resolution.
```{r}
edges = seq(0, 200, 1)
hist(restrictedSet$contb_receipt_amt, edges)
```
I'm not familiar with USA politics, so I can only guess, but I think there are some fixed amounts with which someone can support his/her favorite candidate, like 5, 10, 25, 50, 100 USD. 
```{r}
hist(log(dataSet$contb_receipt_amt))
```