Explore financial contribution of presidential election of California by Sándor Bordé
========================================================

```{r global_options, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(gridExtra)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
finData <- read.csv('support_CA_col.csv')
```

This dataset contains individual donations to the candidates of the Presidential 
Election in 2016. For each data entry, we have the amount of contribution, 
sender, the name of the candidate, date, note, etc.
I have 1,304,346 entry in the dataset.

# Univariate Plots Section

```{r echo=FALSE, Dimension}
dim(finData)
```
```{r echo=FALSE, Variables_list}
str(finData)
```
```{r echo=FALSE, Summary}
summary(finData)
```

The dataset contains 1,304,346 contribution's data; each entry has 18 variables.

There are meaningless or not so informative variables: 

  * *cand_id*: this is redundant if we want to use the name of the candidate too. 
  * *contbr_st*: my dataset contains only data from CA, so it is meaningless.
  * *memo_text*, *receipt_desc*: description of the receipt contains textual data which 
I won't analyze. Besides that, in most cases this field is empty.
  * *memo_cd*: this field only indicates that they provided any memo text.
  * *form_tp*, *file_num*, *tran_id*: these variables serve only administrative purposes, 
 but don't provide any meaningful information.
 
I decided to remove the above variables, so I have ten variables left.

```{r echo=FALSE}
finData = subset(finData, select = -c(cand_id, contbr_st, memo_text, 
                                       receipt_desc, memo_cd, form_tp,
                                       file_num, tran_id))
```

After removing the mentioned features, I got the following ten features.

```{r echo=FALSE, Name_list}
names(finData)
```

My primary interest is the distribution of the contribution amount along 
different properties. After plotting the amount distribution histogram, I found
that there can be outliers which fall far from the most of the values. To see 
the distribution better, I adjusted the axis limits.

```{r}
sum(finData$contb_receipt_amt < 0) / nrow(finData)
```

In my dataset occurred some negative value (which means refund for the 
contributor) what I ignored because I wanted to look at the contribution. 
Negative values made up only the 1.2% of the whole dataset; so I decided to 
remove them from the further analysis.

```{r, Remove_negative}
finData <- finData[(finData$contb_receipt_amt > 0), ]
```

```{r, warning = FALSE}
ggplot(finData, aes((contb_receipt_amt))) +
  geom_histogram(binwidth = 1) +
  xlim(0, 250)
```

As we can see from the initial histogram, supporters tended to give specific 
amount of contribution like \$5, \$10, \$50. But besides this fact, such a small binsize isn't very informative; so I adjusted the histogram further.

```{r, warning = FALSE}
ggplot(finData, aes((contb_receipt_amt))) +
  geom_histogram(binwidth = 0.5) + 
  scale_x_log10()
```

When I transformed the x-axis to log scale, I found that this dataset follows a long-tailed, log-normal distribution.

```{r, warning = FALSE}
ggplot(finData, aes(cand_nm)) +
  geom_bar()
```

# Univariate Analysis

> **Tip**: Now that you've completed your univariate explorations, it's time to
reflect on and summarize what you've found. Use the questions below to help you
gather your observations and add your own if you have other thoughts!

### What is the structure of your dataset?

### What is/are the main feature(s) of interest in your dataset?

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

### Did you create any new variables from existing variables in the dataset?

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?


# Bivariate Plots Section

> **Tip**: Based on what you saw in the univariate plots, what relationships
between variables might be interesting to look at in this section? Don't limit
yourself to relationships between a main output feature and one of the
supporting variables. Try to look at relationships between supporting variables
as well.

```{r echo=FALSE, Bivariate_Plots}

```

# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?


# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!